---
title: "Getting Started with tidyvalidate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with tidyvalidate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction to tidyvalidate

`tidyvalidate` is a package that simplifies data validation in R by providing an intuitive interface to the powerful `validate` package. It helps you ensure data quality before analysis by making it easy to check both column-level and row-level conditions in your data frames.

## Why tidyvalidate?

Data validation in R traditionally faces two challenges:

1. Base R validation functions work well with vectors but become cumbersome when working with data frames, especially for row-level validations
2. The `validate` package offers comprehensive validation capabilities but requires multiple function calls for common use cases

`tidyvalidate` addresses these challenges by providing a streamlined interface that makes data validation both powerful and easy to use.

## Basic Usage

Let's start with a simple example using the built-in `mtcars` dataset:

```{r setup}
library(tidyvalidate)

# Define and run validations
validation_results <- validate_rules(
  mtcars,
  # Column-level validation
  mpg_type = is.numeric(mpg),
  hp_type = is.numeric(hp),
  # Row-level validation
  mpg_threshold = mpg > 15
)
```

### Understanding the Results

The `validate_rules()` function returns a list with two components:

#### 1. Summary of All Validations

```{r}
validation_results$summary
```

The summary table shows:
- Which rules passed or failed
- How many items were checked for each rule
- The number of failures and NA values

#### 2. Detailed Row-Level Errors

```{r}
validation_results$row_level_errors
```

This list contains data tables showing exactly which rows failed each validation rule, but only for row-level rules (like our `mpg_threshold`).

## Advanced Features

### Adding Row Identifiers

When working with real datasets, you often need to identify which specific records failed validation. Here's how to include identifiers in the error output:

```{r}
# Add row names as a column
cars_with_names <- mtcars
cars_with_names$car_name <- rownames(mtcars)
rownames(cars_with_names) <- NULL

# Validate with identifiers
validation_with_ids <- validate_rules(
  cars_with_names,
  mpg_threshold = mpg > 15,
  keep_rl_cols = "car_name"  # Include car_name in error output
)

# View failures with identifiers
validation_with_ids$row_level_errors
```

### Using Dynamic Validation Rules

You can make your validation rules dynamic by using variables. There are two ways to do this:

#### 1. Using Environment Variables

```{r}
# Define threshold in environment
min_mpg <- 12

# Use environment variable in validation
validate_rules(
  mtcars,
  mpg_minimum = mpg > min_mpg
)
```

#### 2. Using the `env` Parameter (Recommended)

```{r}
# Pass variables explicitly
validate_rules(
  mtcars,
  mpg_minimum = mpg > threshold,
  env = list(threshold = 12)  # More explicit and safer
)
```

## Taking Action on Validation Failures

The `action_if_problem()` function helps you handle validation failures appropriately. It offers two modes:

### 1. Stop on Failure

Use this when you want to halt execution if validations fail:

```{r}
try({
  validate_rules(mtcars, mpg_minimum = mpg > 15) |>
    action_if_problem(
      message_text = "Validation failed: Some cars have low MPG",
      problem_action = "stop"
    )
})
```

### 2. Warning on Failure

Use this when you want to continue execution but be notified of failures:

```{r}
results_with_warning <- validate_rules(mtcars, mpg_minimum = mpg > 15) |>
  action_if_problem(
    message_text = "Warning: Some cars have low MPG",
    problem_action = "warning"
  )

# Processing continues and you can still access results
results_with_warning
```

## Best Practices

1. Always include meaningful rule names that describe the validation
2. Use `keep_rl_cols` to include identifying columns in error reports
3. Prefer the `env` parameter over environment variables for dynamic thresholds
4. Choose appropriate actions based on how critical the validation is:
   - Use "stop" for critical data quality issues
   - Use "warning" for advisory checks

## Next Steps

- Explore the `validate` package documentation for more complex validation rules
- Check out the other vignettes for advanced usage patterns
- Consider contributing to the package on GitHub
