---
title: "Introducing tidyvalidate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introducing tidyvalidate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> Make esaer the process of validating data.frames before beginning any process.

## Problem

Start R solution for validations are good for working with vectors, but they aren't so useful when working with data.frames, as we need to apply validations specially at row label.

On the other hand, the `validate` package provides excellent tools to validate data.frames but they divided in several functions to ensure flexibility.

## Solution

`tidyvalidate` wants to simplify the tools provided by the `validate` package to make easier to identify errors and share the results in any useful by like a QC report.

## Example - Finding errors

The `validate_rules()` functions can find errors at column and row level by passing a list with the **summary** and **row_level_errors**.

```{r setup}
library(tidyvalidate)

simple_validation <-
  validate_rules(mtcars,
                 mpg_string = is.character(mpg),
                 hp_numeric = is.numeric(hp),
                 mpg_low = mpg > 15)

```

- **summary** is a data.table showing in the **fails** columns the mistakes found, at column with *mpg_string* and row level with the *mpg_low*.

```{r}
simple_validation$summary
```

- **row_level_errors** is a list of data.tables with a element for each broken rules related to element at row level. In this example, it only shows the *mpg_low* as it is the only at row level with fails rows.

```{r}
simple_validation$row_level_errors
```

### Adding indentifiers at row level

When validating elements at row level is useful to add columns that are not related to the test itself but are useful to identify the individual elements with error.

```{r}
# Creating an unique identifier for each row
mtcars_names <- mtcars
mtcars_names$`Car (Name)` <- rownames(mtcars_names)
rownames(mtcars_names) <- NULL

# Results of validating at row level
validate_rules(mtcars_names,
               mpg_string = is.character(mpg),
               hp_numeric = is.numeric(hp),
               mpg_low = mpg > 15,
               keep_rl_cols = "Car (Name)")$row_level_errors
```


### Validating based on enviroment variables

Some this need to create dynamic validations based on variables from the global environment as you see below.

```{r}
min_mpg <- 12

validate_rules(mtcars_names,
               mpg_string = is.character(mpg),
               hp_numeric = is.numeric(hp),
               mpg_low = mpg > min_mpg,
               keep_rl_cols = "Car (Name)")
```

In those cases, we can also pass a list of elements to the `env` argument.

```{r}
validate_rules(mtcars_names,
               mpg_string = is.character(mpg),
               hp_numeric = is.numeric(hp),
               mpg_low = mpg > var_min_mpg,
               env = list(var_min_mpg = min_mpg),
               keep_rl_cols = "Car (Name)")
```


## Example - Alerting if error

Having a report to identify mistake it's really useful, but we don't always want to see the same summaries if all results are good, but we want be alerted about.

That the main purpose of the `action_if_problem()` function. It has the following actions available:

- **stop**: The *(default)* option will show an error if any the rows listed present an failed element.

```{r}
try({
  validate_rules(mtcars,
                 mpg_string = is.character(mpg),
                 hp_numeric = is.numeric(hp),
                 mpg_low = mpg > 15) |>
    action_if_problem("We shound't have cars with low mpg",
                      problem_action = "stop")
})
```

- **warning**: It will let you know that there is an problem without stop running the code and resturnign the same results provided by the `validate_rules()` function.

```{r}
warning_results <-
  validate_rules(mtcars,
                 mpg_string = is.character(mpg),
                 hp_numeric = is.numeric(hp),
                 mpg_low = mpg > 15) |>
  action_if_problem("We shound't have cars with low mpg",
                    problem_action = "warning")

warning_results
```




