---
title: "Introducing tidyvalidate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introducing tidyvalidate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> Make easier the process of validating data.frames before beginning any process.

## Problem

Standard R solutions for validations are good for working with vectors, but they aren't so useful when working with data.frames, as we need to apply validations specifically at row level.

On the other hand, the `validate` package provides excellent tools to validate data.frames but they are divided into several functions to ensure flexibility.

## Solution

`tidyvalidate` aims to simplify the tools provided by the validate package to make it easier to identify errors and share the results in any useful way, like a QC report.

## Example - Finding errors

The `validate_rules()` function can find errors at column and row level by passing a list with the **summary** and **row_level_errors**.

```{r setup}
library(tidyvalidate)

simple_validation <-
  validate_rules(mtcars,
                 mpg_string = is.character(mpg),
                 hp_numeric = is.numeric(hp),
                 mpg_low = mpg > 15)

```

- **summary** is a data.table showing in the **fails column** the mistakes found, at column with *mpg_string* and row level with the *mpg_low*.

```{r}
simple_validation$summary
```

- **row_level_errors** is a list of data.tables with an element for each broken rule related to elements at row level. In this example, it only shows the *mpg_low* as it is the only row level rule with failed rows.

```{r}
simple_validation$row_level_errors
```

### Adding indentifiers at row level

When validating elements at row level, it is useful to add columns that are not related to the test itself but are useful to identify the individual elements with errors.

```{r}
# Creating an unique identifier for each row
mtcars_names <- mtcars
mtcars_names$`Car (Name)` <- rownames(mtcars_names)
rownames(mtcars_names) <- NULL

# Results of validating at row level
validate_rules(mtcars_names,
               mpg_string = is.character(mpg),
               hp_numeric = is.numeric(hp),
               mpg_low = mpg > 15,
               keep_rl_cols = "Car (Name)")$row_level_errors
```


### Validating based on enviroment variables

Sometimes we need to create dynamic validations based on variables from the global environment as you see below.

```{r}
min_mpg <- 12

validate_rules(mtcars_names,
               mpg_string = is.character(mpg),
               hp_numeric = is.numeric(hp),
               mpg_low = mpg > min_mpg,
               keep_rl_cols = "Car (Name)")
```

In those cases, we can also pass a list of elements to the `env` argument.

```{r}
validate_rules(mtcars_names,
               mpg_string = is.character(mpg),
               hp_numeric = is.numeric(hp),
               mpg_low = mpg > var_min_mpg,
               env = list(var_min_mpg = min_mpg),
               keep_rl_cols = "Car (Name)")
```


## Example - Alerting if error

Having a report to identify mistakes is really useful, but we don't always want to see the same summaries if all results are good, but we want to be alerted about it.

That's the main purpose of the `action_if_problem()` function. It has the following actions available:

- **stop**: The *(default)* option will show an error if any the rows listed present an failed element.

```{r}
try({
  validate_rules(mtcars,
                 mpg_string = is.character(mpg),
                 hp_numeric = is.numeric(hp),
                 mpg_low = mpg > 15) |>
    action_if_problem("We shound't have cars with low mpg",
                      problem_action = "stop")
})
```

- **warning**: It will let you know that there is a problem without stopping the code from running and returning the same results provided by the `validate_rules()` function.

```{r}
warning_results <-
  validate_rules(mtcars,
                 mpg_string = is.character(mpg),
                 hp_numeric = is.numeric(hp),
                 mpg_low = mpg > 15) |>
  action_if_problem("We shound't have cars with low mpg",
                    problem_action = "warning")

warning_results
```
